The Creation of a Programming Language

### 2022-10-29 paper.tex
continue writing paper

\section{Introduction}
The importance of computers in today's world is unquestioned.
But computers are not just the metals and semiconductors they are built from,
but would be unusable without any programs running on them.
Creating such programs, no matter if games, web browsers or even operating systems,
involve the use of programming languages.
Those serve as a bridge between the human mind and its imagination, and
the capabilities of computer hardware and are a huge time-saver when compared
to writing machine-native code, i.e. not using a programming language.

@@

\section{Implementation}
\subsection{Used Software}
At the heart of making any project in the C programming language
stands the C compiler. This project uses
\emph{gcc}, the GNU Compiler Collection, which, as the name suggests, also 
supports more languages than solely C. Substantiating the development process is 
\emph{GNU Make}, a general-purpose build system. It handles tasks like building
the final executable program step-by-step from the project source code and directory clean-up.
The C ecosystem is very easy to use on and fundamentally supported by 
\emph{GNU+Linux}, the operating system used to program on.
The code editor used was \emph{Microsoft Visual Studio Code}, providing great
help with built-in git integration and auto-completion features.
\emph{Git} is the project management and tracking software used to back up and share
the project files on \emph{GitHub.com}, a Microsoft hosted git repository and
web front-end to store and collaborate on source code.
% TODO: add gdb!

@@

The problem to solve in this paper is concisely to make a computer understand
human-written and -readable text, called source code, and follow the instructions
it proposes.
Most interpreters follow a plain scheme to go about this problem, but the
actual implementation of each of the steps can be heavily customized.
Here is the problem, broken down into three steps as commonly done in other
popular programming languages:
\begin{enumerate}\pagelabel{simple_interpreter}
    \item To create meaningful groups of characters from the input string. This
        process is called 'tokenization' and is done with the component called
        tokenizer or lexer. As an example, the string 'aa + bb' could result
        in three tokens, namely 'aa', '+' and 'bb'. In this case, space characters
        are ignored. 
        This is only an example and the grouping of characters could be done 
        in any thinkable, logical way.
    \item To order the tokens into parse-tree, also called syntax-tree, as
        specified by rules. These rules are called a grammar.
        Creating an order is important, as some calculations depend on others, e.g. multiplication
        is always done before addition. An IT data structure that can accommodate
        such requirements is the binary tree. A binary tree consists of tree-nodes
        that each have a maximum of two other tree nodes as their children, and
        a tree node whose child they are, the parent node.
    \item To interpret the parse tree, traversing through it and following the
        instructions each of the tree-nodes represent. The outcome
        will be the result of the computation the program describes.
\end{enumerate}

@@

For the readers who do not know how to read source code, C in particular, here is a short
explanation of the basic syntax of the C language taking the \hrc{src/main.c} file
as an example:
\begin{itemize}
    \item \emph{Preprocessor Instructions.} Lines beginning with hashtags are
        C preprocessor instructions, they are dealt with by the C preprocessor.
        Preprocessing is what the compiler does before actually compiling the
        code.

@@

    \item \emph{Functions.} Every C function has a return type, a
        function name, arguments and a function body in that order.
        Here is an example of a C function:
        \code{pseudocode/function.c}{1}{999}
        This would be an example of a main function, similar to the one
        in \hrc{src/main.c}. It returns a value
        of type 'int', has two arguments, 'argc' and 'argv', and a function body
        which, in the case of the example above, prints the value of the 'argc' argument.

    \item \emph{Pointers.} Pointers are variables that store memory addresses
        of data. In C-like syntax they are defined with the star '*' symbol.
        An example from \hrc{src/main.c} is 'FILE *input;'. That means that input is not
        actually a FILE, but rather a memory address pointing to a spot in memory where a FILE is.
        To dereference the pointer, i.e. to copy the value stored
        at the memory address into a variable on the stack, another star is used like this:
        'FILE value = *input;'.
\end{itemize}

For those who are further interested in the C programming language, there are
countless internet articles or tutorials about reading and understanding or
writing and compiling C code. One such example is the article 'The C Beginner's
Handbook' provided by freeCodeCamp.org linked in the sources
on page \pageref{bibliography}.

Following is a very quick summary of the few 
things happening in the \hrc{src/main.c} file.
First, the \hrc{src/interpreter/interpreter.h} header file is included. 
The main function decides whether to
read a program from the standard input or from a file. 
Then, the interpret function
from the said included header is called with the selected input.
[\cte{freeCodeCamp}]

\subsubsection{The Interpreter}
This section deals with the file 
\hrc{src/interpreter/interpreter.c} which guides
input data through the various stages of the \name{} interpreter as 
seen in figure \ref{f_diagram} on page \pageref{f_diagram}.
Following is a breakdown of the 'interpret' function from said file.

The 'interpret' function calls the parser to generate a syntax-tree from tokens retrieved
from the tokenizer implemented in \hrc{src/interpreter/parser.c} and 
\hrc{src/interpreter/tokenizer.c} respectively.

After that, variables with names are 'localized', meaning their names are
mapped to addresses in memory where their value is stored. Doing this before
bytecode compilation has a performance benefit. 
The mapping is done with a custom hashtable datastructure with average time
complexity of O(log n), n being the amount of different variables.
The hashtable implementation used in \hrc{src/interpreter/memory/hashtable.c}
uses binary search to locate elements variable-sized arrays in the hashtable's buckets.
For collision detection, i.e. to circumvent problems arising when two keys have the
same hash, separate chaining is used. In this approach, linked lists with
key-value pairs are used for each array index and keys are compared directly
to find exact matches.
[\cte{separate_chaining}]

The syntax tree could now be interpreted directly, but as said, is instead converted
to an intermediate internal bytecode format by \hrc{src/interpreter/bytecode.c}.
This step is done so that code that is run multiple times does not convey
the problem of time overhead when traversing the syntax tree, but the current
implementation does not take advantage of this.

The final step is to process or execute said bytecode. This step is performed
by \hrc{src/interpreter/runtime.c}.

@@

Here is an example of tokenization:
An input of 'std.test==23.4' would result in the following tokens according
to the tokenizer implementation of this paper:
\begin{enumerate}
    \item Token of type FIELD with content 'std.text'.
    \item Token of type SYMBOL with content '=='.
    \item Token of type NUMBER with content '23.4'.
\end{enumerate}
FIELD type tokens are later replaced by REFERENCE tokens, effectively mapping
variable names to addresses in memory.
SYMBOL type tokens are mapped to operators such as plus or minux by \hrc{src/implementation/mappings/mapop.c}
during parsing. NUMBER type tokens are replaced by an actual number type, which
is C's 'long double' type as specified in \hrc{src/interpreter/bytecode.h}.

The tokenizer used in this paper is a simple finite-state machine, reading
character by character and determining the token type as it goes.

These tokens are then directly consumed by the parser as described in the next
section.

\subsubsection{The Parser}
The parser, implemented in \hrc{src/interpreter/parser.c}, generates a syntax-tree
from tokens. The algorithm used is inspired by the Pratt Parsing
algorithm first described by Vaughan Pratt in 1973, which is a kind of precedence
parser based on recursive descent. There is an explanation of the algorithm
written by Jonathan Apodaca on dev.io linked in the references on page 
\pageref{bibliography} which this paper's parser takes inspiration from.
[\cte{pratt}, \cte{devio}]

The syntax tree in the implementation of \name{} consists of 'stnode' type
pointers defined in \hrc{src/interpreter/parser.h}. Each such node depending
on its 'nodetype', declared in \hrc{src/interpreter/tokenizer.h},
is either a value-node storing a pointer to data, e.g. a number or text, or a
parent-node referencing two child nodes and an operator, e.g. two numbers as
children and a plus-operator to signify an addition.

The difficult part of implementing a parser is precedence. Precedence together
with associativity define the order of execution of statements. A simple example
would be to perform multiplication before addition, e.g. \emph{2+3*4} would
result in \emph{14} and is not the same as \emph{(2+3)*4}.
Assosiativity is either 'Left to Right' or 'Right to Left'. An example of associativity is
\emph{20/10/2}. Both in C as well as in \name{}, the division operator is
left-to-right associative, meaning the stated expression is the same as
\emph{(20/10)/2} and not as \emph{20/(10/2)}. Both precedence 
and associativity are properties that are defined in the file \hrc{mapop.gpref} for
each operator.

The parser implementation in \hrc{src/interpreter/parser.c} may be difficult to
read and understand, which is why the most important functions are explained here:
\begin{itemize}
    \item \emph{advance.} This function reads a new token from the tokenizer.
    \item \emph{secondary.} The 'secondary' function deals with parsing single,
        simple values that make up the start of an expression, such as numbers,
        fields, strings and it handles brackets, such as indexing with square
        brackets, functions or code blocks with curly braces and sub-expressions
        in standard brackets.
    \item \emph{expr.} The 'expr' function calls 'secondary' which acts
        as the basis for a new expression. It then repeatedly checks
        if an operator follows the secondary token and creates a tree of
        EXPR type stnodes, each having children which are recursively parsed
        and information about the operator.
    \item \emph{parse.} The parse function is a small wrapper which only
        calls the 'expr' function, initiating the parsing process.
\end{itemize}

@@

\subsubsection{The Bytecode Compiler}
The bytecode compiler is an internal part of the interpreter, located in the file
\hrc{src/interpreter/bytecode.c}. Its job is to create a list
of byte-sized instructions and data from the syntax-tree the parser created.
Using such an intermediary bytecode format instead of directly interpreting
the parse-tree can be beneficial. This is the case when the same piece of code
has to be run several times. Instead of generating a time and memory overhead 
having to walk along tree-nodes recursively, only the list of instructions in
the bytecode can be traversed, which takes less time overall.

Not to be confused with machine-native bytecode, the bytecode compiler in \name{}
generates byte-sequences in an internal format that are then interpreted. A
JIT compiler would instead compile the AST down to machine-native bytecode, which
could run directly on the system's hardware and would not have to be interpreted.

To generate the bytecode, \name{} recursively reads through the
binary tree created by the parser and pushes found instructions onto a stack.
The algorithm used is inspired by Edsger Dijkstra's 'Shunting yard 
algorithm', traversing the tree post-order to create
a prefix notation. 
[\cte{shunting_yard}]

A prefix notation is a way to write down mathematical expressions, except
operators precede their operands, in contrast to the more common infix notation,
where operands surround the operator.

In \hrc{src/interpreter/bytecode.c} we can see that first the left operand and then
the right operand are added onto the stack of bytecode. Therefore, when
popping from the stack and evaluating instructions, the operands will have to
be read in reverse.
Here is an example:
\begin{enumerate}
    \item Assume the following mathematical expression in infix notation:
        \expr{\emph{1 - 2 * 4 / (5 + 6)}}
    \item In common prefix notation, also called Polish notation, the
        operators would now precede the operands in this manner:
        \expr{\emph{- 1 (/ (* 2 4) (+ 5 6))}}
    \item But in \name{} the operands are pushed onto the stack is reverse order.
        This would look like the following:
        \expr{\emph{- (/ (+ 6 5) (* 4 2)) 1}}
    \item The brackets are not needed, because for every operator, the amount
        of operands is known. This is the case as e.g. the subtraction operator
        and the negative operator, which are both a minus sign, are distinguished
        beforehand. The above statement would therefore look like this:
        \expr{\emph{- / + 6 5 * 4 2 1}}
        This is how expressions are represented in the bytecode of \name{}.
\end{enumerate}
[\cte{polish_notation}]

@@

The evaluation of expressions also works in a recursive manner. Here is an 
example of \name{}'s bytecode notation and how it is evaluated at runtime:
\begin{itemize}
    \item Assume the input in \name{}'s bytecode to resemble the following expression: 
        \expr{\emph{- 6 * 3 4}}
        In common infix notation, this would be equivalent to 
        \expr{\emph{(4 * 3) - 6}.}
        The question arises, how can an algorithm compute the result of this
        calculation? The answer is recursion.
    \item The runtime will read through the bytecode left-to-right. First it will
        notice the subtraction-operator and accordingly assume that two
        operands must follow.
    \item The function recursively calls itself in order to evaluate the
        right operand and notice the 6, a value. The function simply returns the six
        to the subtraction expression.
    \item Again, the function calls itself to find the next operand.
        As it now reaches a multiplication operator,
        it recursively reads through that sub-expression's right and then left
        operator, which are 3 and 4 respectively.
        The result is computed and returned.
    \item Now the subtraction knows its right- and left-hand operators, 6 and 12.
        The subtraction is performed and the result, 6, is returned.
\end{itemize}

@@

% TODO: add examples in /examples/
\name{} is built upon the implementation of a recursive Pratt
parser and a basic tokenizer. It uses a hash table to map variable names to their
memory addresses. There are no keywords, only a set of non-alphanumeric characters
representing all possible operations. Extending upon that, there are some
standard functions.

\subsection{Syntax}
The syntax of \name{} is unlike most modern, common programming languages' syntax,
but can be compared somone to that of JavaScript.
In \name{}, there are no keywords at all and everything is controlled via operators,
which are non-alphanumeric characters. Such operators include the plus-sign 
for addition, or the equals-sign for variable assignment.

Following is a piece of example \name{} code, printing out the first ten numbers
of the fibonacci sequence, as well as a disection of how that same code is interpreted.
\code{examples/fibonacci.nts}{1}{999}
\begin{itemize}
\item The source code begins by declaring two variables \emph{a} and \emph{b}, which
    will be utilized to keep track of the data used to calculate fibonacci numbers.
    The memory allocation is done by creating a tuple of \emph{a} and \emph{b}, as suggested by 
    the comma seperating the two. Since both variables \emph{a} and \emph{b} are not yet
    known to the interpreter, they are automatically created and assigned a value
    which has no content and the \emph{undefined} typing.
\item The two variables \emph{a} and \emph{b} are then reassigned new values on line
    2 of the program. In this case it is numbers with values zero and one respectively.
    In this example, the two expressions are seperated by a space, but they can
    also be separated by newline-characters or semicolons.
\item Another variable \emph{count} is declared and defined as number in one step on line 4.
    This variable will be used to limit the amount of numbers the program outputs.
\item Yet another variable, \emph{f}, is declared and assigned a function as its
    value on line 6. In \name{}, normal functions do not have any arguments, 
    but they can interact with variables that were created outside their scope.
\item Lines 7 to 10 hold mathematical expressions. At first, a is assigned its current value
    plus the current value of b. Then in line 8, b is assigned the now new value of a minus the
    current value of b, which corresponds to the value of a before it was just reassigned.
    Lastly the variable \emph{count} is decremented by one.
\item Now come the interesting parts of this program.
    On lines 12 and 13, the function \emph{std.print} is referenced. This function is implemented
    in plain C code, which internally is not handled the same as regular functions in \name{},
    and is wrapper to C's function \emph{printf}. The \emph{std.print} function
    takes any \name{} variable or value and prints it to standard output. On line 13, the "\\n"
    is replaced by a newline-character, as \name{} can interpret Java-like escape-sequences.
\item Another standard-library function is \emph{std.if}. It takes two or three arguments.
    If the first argument is true, then the second, else the the third argument or 
    \emph{undefined} is returned. A value is true, only if it is a number not equal to zero
    or a non-empty string. In the case of this example, \emph{std.if} returns \emph{f} if 
    count is greater than zero, and an empty function if count reaches zero. The two
    brackets following the call of the \emph{std.if} function signal yet another call to the
    returned value.
\item The final thing left to do is to get the ball roling by calling the function \emph{f} for
    the first time. As stated above, it will now print numbers and recursively call itsself,
    until the variable \emph{count} reaches a value of zero.
\end{itemize}

\subsection{Name Shadowing}
Here is another example, which in contrast to the example above, focuses on an
original security concept implemented in \name{}.
\code{examples/exec.nts}{1}{999}
The code begins by creating the variable \emph{var}, which holds a string and
\emph{f}, which calls the function \emph{std.exec} with var as argument.
Line number six holds a comment, indicated by the two slashes at the start of the line.
Comments are ignored by the \name{} interpreter and their purpose is to contain notes
and explanations inside the source code itself. Another type of commends are block-comments
which span across multiple lines, such as on lines 9 to 12.
On line seven, the \emph{std.exec} function is called for the first time. It runs the
command it is given as argument in the user's shell, which is most commonly \emph{cmd} on
Microsoft Windows and \emph{bash} on most Linux distributions. In this example, \emph{std.exec}
launches the Firefox browser, but it is capable of all kinds of things that can otherwise be
done in the shell, such as delete files or shut down the computer.

@@

\section{Discussion}

@@

\subsubsection{Numbers}
% TODO

\subsubsection{Memory Leaks}
Memory leaks happen when a program reserves memory to store data, but
never frees up the space for use for other programs, even though the data
has long been forgotten about and will no longer be accessed.
In \name{}, memory leaks are a major problem. Even though barely noticable for small
programs it interprets, \name{} does not have proper setups in place to free up
all used memory that is not used any longer.

Other programming languages, notably object-oriented languages such as Java,
use Garbage-Collection, which automatically tracks unused allocated memory and
frees it up automatically. While this is also the case with \name{}, there are
many internal flaws that have nothing to do with variables used in a \name{}
program, but rather internally.

One reason memory leaks is the nature of the
C programming language, in which the \name{} interpreter was written in. The C language and
compiler do not keep track of all possible states in which certain blocks of memory are
forgotten about and do not warn the developer. Instead, memory leaks happen.
More modern C alternatives such as Rust have an ownership-system in place,
in which any allocated memory is owned by a function. When the function
is done processing, the memory is automatically freed. Such a system
is a helpful tool for developers to create better programs without having to get
lost in the details.
Inventing new ways to make the job of developers easier is what programming
language design is all about.

>>>>

\section{Introduction}
The importance of computers in today's world is unquestioned.
But computers are not just the metals and semiconductors they are built from,
but would be unusable without any software running on them.
Creating programs, no matter if games, web browsers or even operating systems,
involve the use of programming languages.
Those serve as a bridge between the human mind and its imagination, and
the capabilities of computer hardware and are a huge time-saver when compared
to writing machine-native code, i.e. not using a programming language.
By making use of just the term \emph{language} this paper will refer to programming
languages specifically. Sources are mentioned fater cohesive paragraphs in
square brackets, an index can be found in the \emph{Bibliography} on page
\pageref{bibliography}.

@@

\section{Implementation}
\subsection{Used Software}
At the heart of making any project in the C programming language
stands the C compiler. This project uses
\emph{gcc}, the GNU Compiler Collection, which, as the name suggests, also 
supports more languages than solely C. Substantiating the development process is 
\emph{GNU Make}, a general-purpose build system. It handles tasks like building
the final executable program step-by-step from the project source code and directory clean-up.
The C ecosystem is very easy to use on and fundamentally supported by 
\emph{GNU+Linux}, the operating system used to program on.
The code editor used was \emph{Microsoft Visual Studio Code}, providing great
help with built-in git integration and auto-completion features.
The tools \emph{Valgrind} and \emph{gdb}, the GNU Project Debugger were of great help
regarding the finding of errors and memory leaks, see section \ref{memleaks} for details.
\emph{Git} is the project management and version tracking software used to back up and share
the project files on \emph{GitHub.com}, a Microsoft hosted git repository and
web front-end to store and collaborate on source code.

@@

The problem to solve in this paper is concisely to make a computer understand
human-written and -readable text, called source code, and follow the instructions
it proposes.
Most interpreters follow a plain scheme to go about this problem, but the
actual implementation of each of the steps can be heavily customized.
Here is the problem, broken down into three steps as commonly done in other
popular programming languages:
\begin{enumerate}\pagelabel{simple_interpreter}
    \item To create meaningful groups of characters from the input string. This
        process is known as tokenization and is done with the component called
        tokenizer or lexer. As an example, the string \emph{ab + cd} could result
        in three tokens, namely \emph{ab}, \emph{+} and \emph{cd}. In this case, space characters
        are simply ignored and could be left out, unlike in the English language where spaces are relevant. 
        This is only an example and the grouping of characters could be done 
        in any thinkable, logical way.
    \item To order the tokens into parse-tree, also called syntax-tree, as
        specified by rules. These rules are called a grammar.
        Creating an order is important, as some calculations depend on others, e.g. multiplication
        is always done before addition. An IT data structure that can accommodate
        such requirements is the binary tree. A binary tree consists of tree-nodes
        that each have a maximum of two other tree nodes as their children, and
        a tree node whose child they are, the parent node.
    \item To interpret the parse tree, traversing through it and following the
        instructions each of the tree-nodes represent. The outcome
        will be the result of the computation the program describes.
\end{enumerate}

@@

For the readers who do not know how to read source code, C in particular, here is a short
explanation of the basic syntax of the C language taking the \hrc{src/main.c} file
as an example:
\begin{itemize}
    \item \emph{Preprocessor Instructions.} Lines beginning with hashtags are
        C preprocessor instructions, they are dealt with by the C preprocessor.
        Preprocessing is what the compiler does before actually compiling the
        code. %TODO: add src (Preprocessor)

@@

    \item \emph{Functions.} In programming, functions or methods are
        a set of code that accomplish a certain task. They take in data,
        process it and return a result. Calling a function means to evaluate it for a given input.
        In C, every function has a return type, a
        function name, arguments and a function body in that order.
        Here is an example of a C function:
        \code{pseudocode/function.c}{1}{999}
        This would be an example of a main function, similar to the one
        in \hrc{src/main.c}. It returns a value
        of type 'int', has two arguments, 'argc' and 'argv', and a function body
        which, in the case of the example above, prints the value of the 'argc' argument.

    \item \emph{Pointers.} Pointers are variables that store memory addresses. 
        In C syntax they can be defined with the star '*' symbol.
        An example from \hrc{src/main.c} is \expr{FILE *input;.} That means that \emph{input} is not
        actually a FILE, but rather a memory address pointing to a spot in memory where a FILE is.
        To dereference the pointer, i.e. to copy the value stored
        at the memory address into another variable, another star is used like this:
        \expr{FILE value = *input;}.
        \emph{value} now contains a copy of the actual FILE that \emph{input} was pointing to.
\end{itemize}

For those who are further interested in the C programming language, there are
countless internet articles or tutorials about reading and understanding or
writing and compiling C code. One such example is the article 'The C Beginner's
Handbook' provided by freeCodeCamp.org linked in the sources
on page \pageref{bibliography}.
[\cte{freeCodeCamp}]

Following is a very quick summary of the few 
things happening in the \hrc{src/main.c} file.
First, the \hrc{src/interpreter/interpreter.h} header file is included. 
The main function then decides whether to
read a program from a file or from standard input. 
At last, the interpret function
from the said included header is called with the selected input.

\subsubsection{The Interpreter}
This section deals with the file 
\hrc{src/interpreter/interpreter.c} which guides
input data through the various stages of the \name{} interpreter as 
seen in figure \ref{f_diagram} on page \pageref{f_diagram}.
Following is a breakdown of the \emph{interpret} function from said file
as well as an overview of some of the stages that follow.

The \emph{interpret} function calls the parser to generate a syntax-tree from tokens retrieved
from the tokenizer implemented in \hrc{src/interpreter/parser.c} and 
\hrc{src/interpreter/tokenizer.c} respectively.

After that, the names of variables are
mapped to addresses in memory where their value is stored. Doing this before
bytecode compilation has a performance benefit. 
The mapping is acheived with a custom hashtable datastructure with average time
complexity of O(log n), n being the total amount of variables a \name{} program is using.
The hashtable implementation used in \hrc{src/interpreter/memory/hashtable.c}                   % TODO: maybe its own paragraph! interpolation search vs binary search
uses binary search to locate elements variable-sized arrays in the hashtable's buckets.
This is because for such few variables that are used in a standard \name{} program,
an interpolation search algorithm is actually slower.
For collision detection, i.e. to circumvent problems arising when two variable names have the
same hash, a technique called separate chaining is used. In this approach, linked lists with
key-value pairs are used for each array index of the buckets and variable names are compared directly
to find exact matches.
[\cte{separate_chaining}]

The syntax tree from the parser could now be interpreted directly, but as mentioned 
before, is instead converted
to an intermediate internal bytecode format by \hrc{src/interpreter/bytecode.c}.
This step is done so that code that is run multiple times does not have
the problem of time overhead when traversing the syntax tree.

The final step is to process or execute said bytecode. This step is performed
in \hrc{src/interpreter/runtime.c}.

@@

Here is an example of tokenization:
An input of \emph{a.b==23.4} would result in the following tokens according
to the tokenizer implementation of this paper:
\begin{enumerate}
    \item Token of type FIELD with content \emph{a.b}.
    \item Token of type SYMBOL with content \emph{==}.
    \item Token of type NUMBER with content \emph{23.4}.
\end{enumerate}
FIELD type tokens are later replaced by REFERENCE tokens, effectively mapping
variable names to addresses in memory.
SYMBOL type tokens are mapped to operators such as plus or minux by \hrc{src/implementation/mappings/mapop.c}
during parsing. NUMBER type tokens are replaced by an actual number type, which
is C's 'long double' type as specified in \hrc{src/interpreter/bytecode.h}.

While tokenizers can be programmed using regular expressions, 
this paper uses a more simple finite-state machine tokenizer.
It reads character by character, determining the token type as it goes.
It stops when there is a token thad does no longer match the current token type.
Tokens are each requested and integrated into a syntax-tree by the parser as described in the next
section.

\subsubsection{The Parser}
The parser, implemented in \hrc{src/interpreter/parser.c}, generates a syntax tree,
also called parse tree, from tokens. Since this paper's implementation 
does not parse code from a
concise context free grammar, the syntax tree the parser creates is very
similar to an otherwise known abstract syntax tree, AST for short.
ASTs describe source code conceptually and do not contain all syntactical
elements required to parse code.
In this paper the terms parse tree, syntax tree and AST will be used
synonymously.

The algorithm used is inspired by the 'Pratt Parsing
algorithm' first described by Vaughan Pratt in 1973, which is a kind of precedence
parser based on recursive descent. There is an explanation of the algorithm
written by Jonathan Apodaca on \emph{dev.io} linked in the references on page 
\pageref{bibliography} which this paper's parser takes inspiration from.
[\cte{pratt}, \cte{devio}]

The syntax tree in the implementation of \name{} is a binary tree consisting of nodes, internally
called \emph{stnode}. The datatype is defined in the header file \hrc{src/interpreter/parser.h}.
Each node can either be an internal node having one or two nodes as children, or a leaf node
which stands for a value, e.g. a number or a text. Each internal node, also known as parent node, signifies
an operation done on its children. This can be an addition of two numbers for example.

The difficult part of implementing a parser is precedence. Precedence together
with associativity define the order of execution of statements. A simple example
would be to perform multiplication before addition, e.g. \emph{2+3*4} would
result in \emph{14} and is not the same as \emph{(2+3)*4}.
Assosiativity is either 'Left to Right' or 'Right to Left'. An example of associativity is
\emph{20/10/2}. Both in C as well as in \name{}, the division operator is
left-to-right associative, meaning the stated expression is the same as
\emph{(20/10)/2} and not as \emph{20/(10/2)}. Both precedence 
and associativity are properties that are defined in the file \hrc{mapop.gpref} for
each operator.
In the end the parse tree contains all values and 
operations as well as information about the order of execution.

The parser implementation in \hrc{src/interpreter/parser.c} may be difficult to
read and understand, which is why the most important functions are explained here:
\begin{itemize}
    \item \emph{advance.} This function reads the next token from the tokenizer.
    \item \emph{secondary.} This function reads tokens that will become the leaf nodes
        of the parse tree.
        It deals with simple values that make up the start of an expression, such as numbers,
        fields, strings and it handles brackets. 
        In addition, it can handle prefix operators, such as the unary plus or unary minus.
    \item \emph{expr.} The \emph{expr} function calls \emph{secondary} whose result acts
        as the basis for a new expression. It then repeatedly checks
        if an operator follows the secondary token and creates the syntax-tree
        with help of recursive mechanisms. The nature of this function also 
        removees the need for semi-colons,
        which are essential in languages like Java or C.
    \item \emph{parse.} The parse function is a small wrapper which only
        calls the \emph{expr} function a first time, initiating the parsing process.
\end{itemize}

@@

\subsubsection{The Bytecode Compiler}
The bytecode compiler is an internal part of the interpreter, located in the file
\hrc{src/interpreter/bytecode.c}. Its job is to create a list
of byte-sized instructions and data from the syntax-tree the parser created.
Using such an intermediary bytecode format instead of directly interpreting
the parse-tree can be beneficial. This is the case when the same piece of code
has to be run several times. Compared to
having to walk along tree nodes recursively, reading the list of instructions in
the bytecode takes less time overall.

Not to be confused with machine-native bytecode, the bytecode compiler in \name{}
generates byte-sequences in an internal format that are then interpreted. A
Just-In-Time compiler, JIT compiler for short, would instead compile the AST 
down to machine-native bytecode, which
runs directly on the system's hardware and does not have to be interpreted.

To generate the bytecode, \name{} recursively reads through the
binary tree created by the parser and pushes found instructions onto a stack.
The algorithm used is inspired by Edsger Dijkstra's 'Shunting yard 
algorithm', traversing the tree post-order to create
a prefix notation. 
A prefix notation is a way to write down mathematical expressions, except
operators precede their operands, in contrast to the more common infix notation,
where operands surround the operator.
[\cte{shunting_yard}]

In \hrc{src/interpreter/bytecode.c} we can see that first the left operand and then
the right operand are added onto the stack of bytecode. Therefore, when
popping from the stack and evaluating instructions, the operands will have to
be read in reverse.
Here is an example:
\begin{enumerate}
    \item Assume the following mathematical expression in infix notation:
        \expr{\emph{1 - 2 * 4 / (5 + 6)}}
    \item In common prefix notation, also called Polish notation, the
        operators would now precede the operands in this manner:
        \expr{\emph{- 1 (/ (* 2 4) (+ 5 6))}}
    \item But in \name{} the operands are pushed onto the stack is reverse order,
        as popping from the stack again reverses the order back to the original.
        This would look like the following:
        \expr{\emph{- (/ (+ 6 5) (* 4 2)) 1}}
    \item The brackets are not needed, because for every operator, the amount
        of operands is known. This is also the case for the unary minus and unary plus,
        because they are distinguished from their infix counterparts beforehand.
        The above statement would therefore look like this:
        \expr{\emph{- / + 6 5 * 4 2 1}}
        This is how expressions are represented in the bytecode of \name{}.
\end{enumerate}
[\cte{polish_notation}]

@@

The evaluation of expressions works in a recursive manner. Here is an 
example of \name{}'s bytecode notation and how it is evaluated at runtime:
\begin{itemize}
    \item Assume the input in \name{}'s bytecode to resemble the following expression: 
        \expr{\emph{- 6 * 3 4}}
        In common infix notation, this would be equivalent to 
        \expr{\emph{(4 * 3) - 6}.}
        The question arises, how can an algorithm compute the result of this
        calculation? The answer is recursion.
    \item The runtime will read through the bytecode from left to right. First it will
        notice the subtraction operator and accordingly assume that two
        operands must follow.
    \item The function recursively calls itself in order to evaluate the
        right operand and notice the 6, a value and a leaf node in the parse tree. 
        The function simply returns the 6 to the subtraction expression.
    \item Again, the function calls itself to find the next operand.
        As it now reaches a multiplication operator,
        it recursively reads through that sub-expression's right and then left
        operator, which are 3 and 4 respectively.
        The result is computed and returned.
    \item Now the subtraction knows its right- and left-hand operators, 6 and 12.
        The subtraction is performed and the result, 6, is returned.
\end{itemize}

@@

% TODO: add examples in /examples/
\name{} is built upon the implementation of a recursive Pratt
parser and a basic tokenizer. It uses a hash table to map variable names to their
memory addresses. There are no keywords, only a set of non-alphanumeric characters
representing all possible operations. Extending upon that, there are some
standard functions that enlargen the capabilities of the language.

\subsection{Syntax}
The syntax of \name{} is unlike most modern, common programming languages' syntax,
but comes closest to that of JavaScript.
In \name{}, there are no keywords at all and everything is controlled via operators,
which are non-alphanumeric characters. Such operators include the plus-sign 
for addition, the equals-sign for variable assignment or a set of brackets to signal
a function call.

Following is a piece of example \name{} code, printing out the first ten numbers
of the fibonacci sequence, as well as a disection of how that same code is interpreted.
\code{examples/fibonacci.nts}{1}{999}
\begin{itemize}
\item The source code begins with the definition two variables \emph{a} and \emph{b}, which
    will be utilized to keep track of the data used to calculate fibonacci numbers.
    They are each assigned a number, zero and one respectively.
    Since the variable names \emph{a} and \emph{b} are not yet
    known to the interpreter, they are automatically allocated in the current scope
    and assigned a default value. That default value is then overwritten with the
    assignment operator.
\item Another variable \emph{count} is declared and defined as the number ten on line 4.
    This variable will be used to limit the amount of numbers the program outputs.
\item Now the function \emph{std.repeat} is called. This function is a
    standard library function, meaning it is included in the \name{} interpreter and
    written in C. Another function is passed as argument to the function call.
    This function will be repeated.
\item On line 7 and 8, some calculations are done.
    Remember, the equals sign stands for an assignment, not an
    equation. First, \emph{a} is assigned the sum of \emph{a} and \emph{b}. Then 
    \emph{b} is set to be the new \emph{a} minus \emph{b}, 
    which is in turn the same as \emph{a} before the reassignment on line 7.
\item On lines 10 and 11, the standard library function \emph{std.print} is used
    to print out the value of a and a line break. The backslash in the text
    that is print out signifies a so-called escape-sequence. The \emph{n} stands
    for a newline, but other letters such as a \emph{t} for a tabulator are possible.
\item Lastly, line 13 contains another calculation in which the variable \emph{count}
    is decremented by one. Since this is the last instruction in the inner function,
    it is simoultaneously its return value, e.g. the value of count is returned
    to the \emph{std.repeat} function. The \emph{std.repeat} function will
    repeat the inner function until the returned value is zero or undefined.
    This means the inner function is executed exacly ten times, because
    the initial value of count has to be decremented by one ten times until
    it equals zero.
\end{itemize}

\subsection{Name Shadowing}
Here is another example, which in contrast to the example above, focuses on an
original security concept implemented in \name{}.
\code{examples/exec.nts}{1}{999}
The code begins by creating the variable \emph{var}, which holds a string and
\emph{f}, which calls the function \emph{std.exec} with var as argument.
Line number seven holds a comment, indicated by the two slashes at the start of the line.
Comments are ignored by the \name{} interpreter and their purpose is to contain notes
and explanations inside the source code itself. Another type of commends are block-comments
which span across multiple lines, such as on lines 9 to 12.
On line seven, the \emph{std.exec} function is called for the first time. It runs the
command it is given as argument in the user's shell, which is most commonly \emph{cmd} on
Microsoft Windows and \emph{bash} on most Linux distributions. In this example, \emph{std.exec}
launches the Firefox browser, but it is capable of all kinds of things that can otherwise be
done in the shell, such as delete files or shut down the computer.

@@

% TODO: link more examples

\section{Discussion}
The following paragraphs will go over some positive results and some caveats of the
final product.

The name \name{} was chosen, because it implised this programming language
to be a learning project as well as a scripting language, which employ
high-level abstractions of underlying operations and interpret one line
at a time. The file name extension \emph{.nts} is used in the examples. This
file extension is not required, but was chosen as it stands for \emph{Not TypeScript}, because
the language neither enforces strict typing, nor is it Microsoft's \emph{TypeScript}.

@@

\subsubsection{Numbers}
Another goal that turned out not to be practical to acheive was to use just as much
memory for each variable as needed and not more. This means every variable would
have to be assigned a datatype, containing information about the size and possible values.
Taking C as a reference, the datatype \emph{char} is exactly one byte in size, and
can therefore be in exactly 2^8 or 256 different states. But in C, information about
the datatype can be discarded at compile time, whereas in an interpreter like \name{} such
information needs to persist throughout the program's lifetime.

\subsubsection{Memory Leaks}\label{memleaks}
Memory leaks happen when a program reserves memory to store data, but
never frees up the space for use for other programs, even though the data
has long been forgotten about and will no longer be accessed.
In \name{}, memory leaks are a major problem. Even though barely noticable in small
programs it interprets, \name{} does not have proper setups in place to free up
all used memory that is not used any longer.

Manx rogramming languages, such as Java and Python,
make use of Garbage-Collection, a strategy to track unused allocated memory and
free it up automatically. While \name{} also tries to make use of this technique,
there are many flaws introduced while developing the interpreter that hinder 
Garbage-Collection from working correctly.
